---
title: "coursework_Q3"
output: html_document
date: "2023-03-17"
---

# Setting up the workspace
## Installing the necessary packages
```{r}
# install.packages("DBI")
library(DBI)
# install.packages("dplyr")
library(dplyr)
# install.packages("ggplot2")
library(ggplot2)
# install.packages("skimr")
library(skimr)
```

## Setting and checking working directory
```{r}
# Setting the working directory
setwd("~/Library/CloudStorage/OneDrive-SIM-SingaporeInstituteofManagement/Year 2/ST2195 - Programming for Data Science/Coursework/dataverse_files")
getwd()
```

```{r}
# Connecting to the database
conn <- dbConnect(RSQLite::SQLite(), "coursework_r.db")
```

# Data Wrangling
```{r}
# Checking and dealing with anomalies in data set
airports_check <- dbGetQuery(conn, "SELECT iata FROM airports")
```

```{r}
skim(airports_check)
```

# Question 3
## 3. How does the number of people flying between different locations change over time?
### Part 1: Top air traffic volume for airports
```{r}
# Identifying the top 10 air traffic volume for airports
q3_airport_traffic <- dbGetQuery(conn,
  "SELECT airports.airport, airports.iata, COUNT(*)
   FROM ontime JOIN airports ON ontime.Dest = airports.iata
   WHERE Cancelled = 0 AND Diverted = 0
   GROUP BY airports.airport
   ORDER BY COUNT(*) DESC LIMIT 10
  ")
```

```{r}
# Changing the column names of q3_airport_traffic
colnames(q3_airport_traffic) <- c("Airport", "IATA Code", "Number of Flights")

# Printing the output of q3_airport_traffic
q3_airport_traffic

# Checking the number of flight records
sum(q3_airport_traffic$`Number of Flights`) # 7,124,177 flights from the top 10 airports
```

```{r}
# Obtaining data needed to visualize the flight volumes for the top 10 airports
q3p1 <- dbGetQuery(conn,
  "SELECT airports.airport, airports.iata, COUNT(*) AS NumOfFlights, ontime.Year, ontime.Month
   FROM ontime JOIN airports ON ontime.Dest = airports.iata
   WHERE Cancelled = 0 AND Diverted = 0
   GROUP BY airports.airport, ontime.Year, ontime.Month
   ORDER BY airports.airport, ontime.Year, ontime.Month
  ")
```


```{r}
# Changing the column names of q3p1
colnames(q3p1) <- c("Airport", "IATA Code", "Number of Flights", "Year", "Month")

# Checking that the column names were changed
head(q3p1)
```

```{r}
# Create a new variable that combines the Year and Month columns
q3p1$YearMonth <- as.Date(paste(q3p1$Year, sprintf("%02d", q3p1$Month), "01", sep = "-"), format = "%Y-%m-%d")

# To check the total number of flights
sum(q3p1$`Number of Flights`) # 21,271,928 flights
```

```{r}
# Obtaining the number of flights for the top 10 airports
top_airports <- q3p1 %>%
  group_by(Airport) %>%
  summarize(total_flights = sum(`Number of Flights`)) %>%
  arrange(desc(total_flights)) %>%
  slice(1:10) %>%
  pull(Airport)
print(top_airports)

q3p1_top10 <- q3p1 %>% filter(Airport %in% top_airports)

# Number of flights of the top 10 airports
sum(q3p1_top10$`Number of Flights`) # 7124177 flights
```

```{r}
# Create a line chart of the top 10 airports from 2005 to 2007
ggplot(q3p1_top10, aes(x = YearMonth, y = `Number of Flights`, 
                       color = `IATA Code`)) + 
  geom_line() + 
  scale_x_date(date_labels = "%b-%Y", date_breaks = "2 months") +
  labs(title = "Figure 3a: Air Traffic Volume Trends of Top 10 US Airports", 
       x = "Time Period", y = "Number of Flights", color = "Airports") +
  theme_bw() +
  theme(plot.title = element_text(size = 13, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 12), 
        axis.text.x = element_text(size = 9, angle = 45, vjust = .5, hjust = .5),
        axis.text.y = element_text(size = 11)) +
  ylim(10000, 40000)
```

#### Part 2: Top 10 air traffic volume for states
```{r}
# Identifying the top 10 air traffic volume for states
q3_state_traffic <- dbGetQuery(conn,
  "SELECT airports.state, COUNT(*)
   FROM ontime JOIN airports ON ontime.Dest = airports.iata
   WHERE Cancelled = 0 AND Diverted = 0
   GROUP BY airports.state
   ORDER BY COUNT(*) DESC LIMIT 10
  ")
```

```{r}
# Changing the column names of q3_state_traffic
colnames(q3_state_traffic) <- c("State", "Number of Flights")

# Printing the output of q3_state_traffic
q3_state_traffic

# Checking the total number of flights
sum(q3_state_traffic$`Number of Flights`) # 12,697,714 flights from the top 10 states
```

```{r}
# Obtaining data needed to visualize the flight volumes for the top 10 states
q3p2 <- dbGetQuery(conn,
  "SELECT ontime.Year, ontime.Month, airports.state, COUNT(*) AS NumOfFlights
   FROM ontime JOIN airports ON ontime.Dest = airports.iata
   WHERE Cancelled = 0 AND Diverted = 0 AND airports.state IS NOT NULL
   GROUP BY airports.state, ontime.Year, ontime.Month
   ORDER BY airports.state, ontime.Year, ontime.Month
  ")
```

```{r}
# Changing the column names of q3p2 
colnames(q3p2) <- c("Year", "Month", "State", "Number of Flights")

# Checking that the column names were changed
head(q3p2)
```

```{r}
# Create a new variable that combines the Year and Month columns
q3p2$YearMonth <- as.Date(paste(q3p2$Year, sprintf("%02d", q3p2$Month), "01", sep = "-"), format = "%Y-%m-%d")

# To check the total number of flights
sum(q3p2$`Number of Flights`) # 21,257,880 flights
```

```{r}
# Obtaining the number of flights for the top 10 states
top_states <- q3p2 %>%
  group_by(State) %>%
  summarize(total_flights = sum(`Number of Flights`)) %>%
  arrange(desc(total_flights)) %>%
  slice(1:10) %>%
  pull(State)

q3p2_top10 <- q3p2 %>% filter(State %in% top_states)

# Number of flights of the top 10 states
sum(q3p2_top10$`Number of Flights`) # 12,697,714 flights
```

```{r}
# Create a line chart of the top 10 states from 2005 to 2007
ggplot(q3p2_top10, aes(x = YearMonth, y = `Number of Flights`, color = `State`)) + 
  geom_line() + 
  scale_x_date(date_labels = "%b-%Y", date_breaks = "2 months") +
  labs(title = "Figure 3b: Air Traffic Volume Trends of Top 10 US States", 
       x = "Time Period", y = "Number of Flights", color = "States") +
  theme_bw() +
  theme(plot.title = element_text(size = 13, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 12), 
        axis.text.x = element_text(size = 9, angle = 45, vjust = .5, hjust = .5),
        axis.text.y = element_text(size = 11)) +
  ylim(14000, 80000)
```

#### Part 3: Top 10 air traffic volume for cities
```{r}
# Identifying the top 10 air traffic volume for cities
q3_city_traffic <- dbGetQuery(conn,
  "SELECT airports.city, COUNT(*)
   FROM ontime JOIN airports ON ontime.Dest = airports.iata
   WHERE Cancelled = 0 AND Diverted = 0
   GROUP BY airports.city
   ORDER BY COUNT(*) DESC LIMIT 10
  ")
```

```{r}
# Changing the column names of q3_city_traffic
colnames(q3_city_traffic) <- c("City", "Number of Flights")

# Printing the output of q3_city_traffic
q3_city_traffic

# Number of flights of the top 10 cities
sum(q3_city_traffic$`Number of Flights`) # 7,818,747 flights from the top 10 cities
```

```{r}
# Obtaining data needed to visualize the flight volumes for the top 10 cities
q3p3 <- dbGetQuery(conn,
  "SELECT ontime.Year, ontime.Month, airports.city, COUNT(*) AS NumOfFlights
   FROM ontime JOIN airports ON ontime.Dest = airports.iata
   WHERE Cancelled = 0 AND Diverted = 0 AND airports.city IS NOT NULL
   GROUP BY airports.city, ontime.Year, ontime.Month
   ORDER BY airports.city, ontime.Year, ontime.Month
  ")
```

```{r}
# Changing the column names of q3
colnames(q3p3) <- c("Year", "Month", "City", "Number of Flights")

# Checking that the column names were changed
head(q3p3)
```

```{r}
# Create a new variable that combines the Year and Month columns
q3p3$YearMonth <- as.Date(paste(q3p3$Year, sprintf("%02d", q3p3$Month), "01", sep = "-"), format = "%Y-%m-%d")

# To check the total number of flights
sum(q3p3$`Number of Flights`) # 21,257,880 flights
```

```{r}
# Obtaining the number of flights for the top 10 cities
top_cities <- q3p3 %>%
  group_by(City) %>%
  summarize(total_flights = sum(`Number of Flights`)) %>%
  arrange(desc(total_flights)) %>%
  slice(1:10) %>%
  pull(City)

q3p3_top10 <- q3p3 %>% filter(City %in% top_cities)

# Number of flights of the top 10 cities
sum(q3p3_top10$`Number of Flights`) # 7,818,747 flights
```

```{r}
# Create a line chart of the top 10 cities from 2005 to 2007
ggplot(q3p3_top10, aes(x = YearMonth, y = `Number of Flights`, color = `City`)) + 
  geom_line() + 
  scale_x_date(date_labels = "%b-%Y", date_breaks = "2 months") +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(title = "Figure 3c: Air Traffic Volume Trends of Top 10 US Cities", 
       x = "Time Period", y = "Number of Flights", color = "Cities") +
  theme_bw() +
  theme(plot.title = element_text(size = 13, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 12), 
        axis.text.x = element_text(size = 9, angle = 45, vjust = .5, hjust = .5),
        axis.text.y = element_text(size = 11))
```

#### Part 4: Top 10 air traffic volume for flight routes
```{r}
# Identifying the top 10 air traffic volume for flight routes
q3_route_traffic <- dbGetQuery(conn,
   "SELECT ontime.Origin, ontime.Dest, COUNT(*) AS NumOfFlights
    FROM ontime
    WHERE Cancelled = 0 AND Diverted = 0
    GROUP BY ontime.Origin, ontime.Dest
    ORDER BY NumOfFlights DESC
    LIMIT 10
   ")
```

```{r}
# Changing the column names of q3_route_traffic
colnames(q3_route_traffic) <- c("Origin", "Destination", "Number of Flights")

# Printing the output of q3_route_traffic
q3_route_traffic

# Number of flights of the top 10 routes
sum(q3_route_traffic$`Number of Flights`) # 375,850 flights from the top 10 flight routes
```

```{r}
# Adding a new column named "Route"
dbExecute(conn, "ALTER TABLE ontime ADD Route")
dbListFields(conn, "ontime")
```

```{r}
# Grouping flight routes based on their origin and destination
dbExecute(conn, "UPDATE ontime SET Route = Origin || '-' || Dest")
```

```{r}
# Obtaining data needed to visualize the flight volumes for the top 10 flight routes
q3p4 <- dbGetQuery(conn,
                   "SELECT ontime.Year, ontime.Month, Route, COUNT(*) AS NumOfFlights
                   FROM ontime
                   WHERE Cancelled = 0 AND Diverted = 0 AND Route IS NOT NULL
                   GROUP BY Route, Year, Month
                   ORDER BY Route, Year, Month
                   ")
```

```{r}
# Changing the column names of q3p4
colnames(q3p4) <- c("Year", "Month", "Route", "Number of Flights")

# Checking that the column names were changed
head(q3p4)
```

```{r}
# Create a new variable that combines the Year and Month columns
q3p4$YearMonth <- as.Date(paste(q3p4$Year, sprintf("%02d", q3p4$Month), "01", sep = "-"), format = "%Y-%m-%d")

# To check the total number of flights
sum(q3p4$`Number of Flights`) # 21,271,928 flights
```

```{r}
# Obtaining the number of flights for the top 10 flight routes
top_routes <- q3p4 %>%
  group_by(Route) %>%
  summarize(total_flights = sum(`Number of Flights`)) %>%
  arrange(desc(total_flights)) %>%
  slice(1:10) %>%
  pull(Route)

q3p4_top10 <- q3p4 %>% filter(Route %in% top_routes)

# Number of flights of the top 10 flight routes
sum(q3p4_top10$`Number of Flights`) # 375,850 flights
```

```{r}
# Create a line chart of the top 10 flight routes from 2005 to 2007
ggplot(q3p4_top10, aes(x = YearMonth, y = `Number of Flights`, color = `Route`)) + 
  geom_line() + 
  scale_x_date(date_labels = "%b-%Y", date_breaks = "2 months") +
  labs(title = "Figure 3d: Air Traffic Volume Trends of Top 10 US Flight Routes", 
       x = "Time Period", y = "Number of Flights", color = "Routes") +
  theme_bw() +
  theme(plot.title = element_text(size = 13, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 12), 
        axis.text.x = element_text(size = 9, angle = 45, vjust = .5, hjust = .5),
        axis.text.y = element_text(size = 11)) +
  ylim(500, 1930)
```

```{r}
# Disconnecting from the database
dbDisconnect(conn)
```